Rates of true polar wander in convecting planets

### 2015-04-02 tpw_rate.tex
Disorganized work

\begin{equation}
\begin{aligned}
 \rho{\bf \Omega \times \Omega \times r}= - \nabla P + \nabla^2 {\bf u} + \rho {\bf g}
\label{navier_stokes}
\end{aligned}
\end{equation}

@@

In particular, the most important numbers are the Rayleigh number, characterizing the vigor of convection, and the ratio of centrifugal to gravitational forces.
This second nondimensional number does not have a uniformly agreed-upon name, but has been called a Froude number in analogy with other applications of inertial-to-gravitational effects \citep{mckenzie1968influence}, which we adopt here.
Since we have begun with equations that do not have inertia or compressibility, we have implicitly thrown out the dependence on the nondimensional numbers which characterize those effects (e.g., the Prandtl and dissipation numbers).
It would be straightforward to include them, but does not affect the overall treatment of the effects of rotation.

@@

Nondimensionalizing with these parameters we obtain
\begin{equation}
\begin{aligned}
 - \nabla P^* + \; \nabla^2{\bf v} - \mathrm{Ra} \; T \; {\bf g} + \mathrm{Ra \; Fr}\; T \;{\mitbf{\omega} \times \mitbf{\omega} \times r} = 0
\end{aligned}
\end{equation}

where we have introduced a dynamic pressure $P^* = P - P_0$.

@@

This can be transformed into an angular momentum equation by crossing with $\bf r$ and integrating over the volume of the mantle:

\begin{equation}
-\int_V {\bf r} \times \nabla P^* + \int_V {\bf r} \times \nabla^2 {\bf v} - \mathrm{Ra} \int_V T {\bf r \times g} + \mathrm{Ra \; Fr} \int_V T {\bf r \times \mitbf{\omega} \times \mitbf{\omega} \times r} = 0
\end{equation}

\begin{equation}
\begin{aligned}
\int_V T {\bf r \times \mitbf{\omega} \times \mitbf{\omega} \times r} = 0 \\
\end{aligned}
\end{equation}

@@

The time constant $T_1$ is related to the viscous tidal relaxation time of the planet. It is frequently represented as a weighted average of the different relaxation modes of the planet \citep[e.g.][]{ricard1993polar, greff2004upwelling}, but for our purposes it is enough to approximate it as a single mode:
\begin{equation}
T_1 \sim \eta / \rho g R = \frac{R^2}{\kappa} \Gamma/\mathrm{Ra}
\end{equation}

>>>>

\begin{equation}
\begin{aligned}
- \nabla P + \eta \nabla^2 {\bf g} =  \rho {\bf g} -  \rho{\bf \Omega \times \Omega \times r}
\label{navier_stokes}
\end{aligned}
\end{equation}

@@

In particular, the most important numbers are the Rayleigh number, characterizing the vigor of convection, and the ratio of centrifugal to gravitational forces.
This second nondimensional number does not have a uniformly agreed-upon name, but has been called a Froude number in analogy with other applications of inertial-to-gravitational effects \citep{mckenzie1968influence}, which we adopt here.
Since we have begun with equations that do not have inertia or compressibility, we have implicitly thrown out the dependence on the nondimensional numbers which characterize those effects (e.g., the Prandtl and dissipation numbers).
It would be straightforward to include them, but does not affect the overall treatment of this scaling.

Customarily a hydrostatic reference state is defined as the solution to \ref{navier_stokes} where there is no flow:

\begin{equation}
\begin{aligned}
- \nabla P_0 =  \rho_0 {\bf g} -  \rho_0 {\bf \Omega \times \Omega \times r}
\label{hydrostatic}
\end{aligned}
\end{equation}

where $P_0$ is the reference pressure.
We can define a ``dynamic pressure'' $P^* = P - P_0$.

Nondimensionalizing with the parameters in Table \ref{nondim} we find
\begin{equation}
\begin{aligned}
 - \nabla P^* + \; \nabla^2{\bf v} - \mathrm{Ra} \; T \; {\bf g} + \mathrm{Ra \; Fr}\; T \;{\mitbf{\omega} \times \mitbf{\omega} \times {\bf r}} = 0
\end{aligned}
\end{equation}

The figure for the planet which is consistent with the hydrostatic state we call $V_H$.
The actual figure of the planet's mantle, $V$, may have some disequilibrium with this hydrostatic shape: $V = V_H + \Delta V $.
The introduction of $\Delta V$ requires another nondimensional number to characterize it, and we find that the quantity $\Gamma = \alpha \Delta T$ is convenient.

Equation \ref{navier_stokes} can be transformed into an angular momentum equation by crossing with $\bf r$ and integrating over the volume of the mantle:

\begin{equation}
\begin{aligned}
-\int_V {\bf r} \times \nabla P + \int_V \eta {\bf r} \times \nabla^2 {\bf u} - \int_V \rho {\bf r \times g} + \int_V \rho {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} &= 0 \\
\end{aligned}
\end{equation}

\begin{equation}
\begin{aligned}
\int_V \rho {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} &= 0 \\
\end{aligned}
\end{equation}

which is simply a statement that a quasistatic body will rotate around the principle axis of its total moment of inertia.
Splitting the volume into $V_H$ and $\Delta V$ and substituting \ref{eos} we find

\begin{equation}
\begin{aligned}
&\int_{V_H} \rho_0 {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} + 
\int_{V_H} \rho_0 \alpha T {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} +  \\
&\int_{\Delta V} \rho_0 {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} + 
\int_{\Delta V} \rho_0 \alpha T {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} = 0  \\
\end{aligned}
\end{equation}

The first term of this equation is zero due to the hydrostatic equation.  
The fourth term is negligible due to being second order in the smallness parameters $\Delta V/V_H$ and $\alpha \Delta T$.

@@

@@

@@

The time constant $T_1$ is a viscous tidal relaxation time of the planetary mantle. It is frequently represented as a weighted average of the different relaxation modes \citep[e.g.][]{ricard1993polar, greff2004upwelling}, but for our purposes it is enough to approximate it as a single mode:
\begin{equation}
T_1 \sim \eta / \rho g R = \frac{R^2}{\kappa} \Gamma/\mathrm{Ra}
\end{equation}
